Server-Side Probabilistic Programming

### 2019-04-15 paper/paper.tex
in progress, overall structure - fill it!

\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage{url}
\usepackage{hyperref}

\linespread{1.03}

\acmConference[LAFI'2019]{Languages for Inference}{January 15,
2019}{Lisbon, Portugal}

\setcopyright{none}

@@

\begin{abstract}
	We share our experiences from integrating probabilistic
	programming into a server-side software system and
	present our ongoing work on probabilistic programming
	facility for Go, a modern programming language of choice for
	server-side software development. We demonstrate how a
	lightweight efficient probabilistic programming facility can
	be added to an existing programming language.  Server-side
	application of probabilistic programming poses challenges
	for a probabilistic programming system. We discuss the
	challenges and our experience in overcoming them, and
	suggest guidelines that can help in a wider adoption of
	probabilistic programming in server-side software systems.
\end{abstract}

@@

\begin{sloppypar}
\section{Challenges of Server-Side Probabilistic Programming}

@@

\section{Probabilistic Programming Facility for Go}

@@

In line with the guidelines, we have implemented a probabilistic
programming facility for the Go programming language,
\texttt{infergo} (\url{http://infergo.ml/}). We have chosen Go
because Go is a small but expressive programming language with
efficient implementation, which has recently become quite
popular for computation-intensive server-side programming. At
the time of writing, the implementation is still at an early
stage. However, this facility is already used in production
environment for inference of mission-critical algorithm
parameters.

A probabilistic model in \texttt{infergo} is an implementation
of the \texttt{Model} interface requiring a single method
\texttt{Observe} which accepts a vector (a Go \textit{slice}) of
floats, the parameters to infer, and returns a single float,
interpreted as unnormalized log-likelihood of the posterior
distribution. Implementation of model methods can be written
in virtually unrestricted Go and use any Go libraries.

For inference, \texttt{infergo} relies on automatic
differentiation. The source code of the model is
translated by a command-line tool provided by \texttt{infergo}
into an equivalent model with reverse-mode automatic
differentiation of the log-likelihood with respect 
to the parameters applied. The differentiation operates
on the built-in floating-point type and incurs only a small
computational overhead. However, even this overhead is avoided
when the model code is executed outside of inference algorithms:
both the original and the differentiated model are
simultaneously available to the rest of the program code, so
the methods can be called on the differentiated model for
inference, and on the original model for the most efficient
execution with the inferred parameters.

The Go programming language and development environment offer
capabilities which made implementation of \texttt{infergo}
affordable.

@@

Table~\ref{tab:memory-runtime} provides memory consumption and
running time measurements on basic models to
illustrate  \texttt{infergo}'s performance.
The measurements were obtained on a 2.3GHz Intel Core 5
CPU with 8GB of memory for 1000 iterations of Hamiltonian Monte
Carlo with 10 leapfrog steps. Note that log-likelihood computation 
for standard distributions is not optimized yet. Quite the
opposite: since models in \texttt{infergo} are fully composable,
primitive distributions are themselves implemented as
\texttt{infergo} models and automatically differentiated.

@@

A lightweight probabilistic programming facility similar to
\texttt{infergo} can be added to most modern general-purpose
programming languages, in particular those used in implementing
large-scale software systems, making probabilistic
programming inference more accessible in server-side
applications.
\end{sloppypar}

>>>>

\documentclass[sigplan,review,10pt,anonymous]{acmart}
\usepackage{url}
\usepackage{hyperref}

\acmConference[Onward! 2019]{SPLASH Onward! 2019}{October 20-25, 2019}{Athens, Greece}

% Copyright
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

@@

\begin{abstract}
	We share our experiences from integrating probabilistic
	programming into a server-side software system and introduce
	a probabilistic programming facility for Go, a modern
	programming language of choice for server-side software
	development. We demonstrate how a lightweight but powerful
	and efficient probabilistic programming facility can be
	added to an existing programming language.  Server-side
	application of probabilistic programming poses challenges
	for a probabilistic programming system. We discuss the
	challenges and our experience in overcoming them, and
	suggest guidelines that can help in a wider adoption of
	probabilistic programming in server-side software systems.
\end{abstract}

@@

\begin{sloppypar}

\section{Introduction}

The promise of probabilistic programming.

The language war.

A solution: probabilistic programming without a DSL.

Organization of the paper.

\subsection*{Contributions}

\section{Related Work}

\section{Challenges of Server-Side Probabilistic Programming}

@@

\section{Guidelines to Implementation of a Probabilistic Programming Facility}

@@

\section{Probabilistic Programming in Go}

In line with the guidelines, we have implemented a probabilistic
programming facility for the Go programming language,
Infergo. %(\url{http://infergo.ml/}).
We have chosen Go because Go is a small but expressive
programming language with efficient implementation, which has
recently become quite popular for computation-intensive
server-side programming.  the time of writing, the
implementation is still at an early stage.  This facility
ialready used in production environment for inference of
mission-critical algorithm parameters.

\subsection{An Overview of Infergo}

A probabilistic model in Infergo is an implementation of an
interface requiring a single method \texttt{Observe} which
accepts a vector (a Go \textit{slice}) of floats, the parameters
to infer, and returns a single float, interpreted as
unnormalized log-likelihood of the posterior distribution.
Implementation of model methods can be written in virtually
unrestricted Go and use any Go libraries.

For inference, Infergo relies on automatic differentiation. The
source code of the model is translated by a command-line tool
provided by Infergo into an equivalent model with reverse-mode
automatic differentiation of the log-likelihood with respect to
the parameters applied. The differentiation operates on the
built-in floating-point type and incurs only a small
computational overhead. However, even this overhead is avoided
when the model code is executed outside of inference algorithms:
both the original and the differentiated model are
simultaneously available to the rest of the program code, so the
methods can be called on the differentiated model for inference,
and on the original model for the most efficient execution with
the inferred parameters.

The Go programming language and development environment offer
capabilities which made implementation of Infergo affordable.

@@

\subsection{Probabilistic ``Hello, World!''}

TODO: Example

\section{Model Syntax and Rules of Differentiation}

\section{Design and Implementation Highlights}

\subsection{Choice of Programming Language}

\subsection{Automatic Differentiation}

\subsection{Inference}

\subsection{Composability of Models}

\subsection{Stochasticity and Streaming}

\section{Case Studies}

\subsection{Performance}

Table~\ref{tab:memory-runtime} provides memory consumption and
running time measurements on basic models to illustrate
Infergo's performance.  The measurements were obtained on a
2.3GHz Intel Core 5 CPU with 8GB of memory for 1000 iterations
of Hamiltonian Monte Carlo with 10 leapfrog steps. Note that
log-likelihood computation for standard distributions is not
optimized yet. Quite the opposite: since models in Infergo are
fully composable, primitive distributions are themselves
implemented as Infergo models and automatically differentiated.

@@

\section{Conclusion}

A lightweight probabilistic programming facility similar to
Infergo can be added to most modern general-purpose programming
languages, in particular those used in implementing large-scale
software systems, making probabilistic programming inference
more accessible in server-side applications.

\section*{Acknowledgements}

\end{sloppypar}
